---
title: "MTT Assay Analysis"
author: "Victoria Flanary"
date: "2025-01-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objective
Analyze results from the MTT assay experiments of neuroblastoma cell lines
treated with 3 different concentrations of 13cRA at 3 different timepoints.
Experiments were conducted by Ian Doty and results were analyzed by Victoria
Flanary.

# Set-up
```{r load packages}
library(tidyverse)
library(ggpubr)
library(rstudioapi)
library(here)
library(styler)
library(lintr)
```

```{r load the results}
mtt_results <- read_csv(
  here("MTT_Assay", "13cRA_MTT_Assay_Results.csv")
)
```

```{r format columns}
mtt_results$cell_line <- factor(
  mtt_results$cell_line,
  levels = unique(mtt_results$cell_line)
)

mtt_results$timepoint <- factor(
  mtt_results$timepoint,
  levels = unique(mtt_results$timepoint)
)

mtt_results$date <- as.factor(mtt_results$date)

mtt_results$concentration <- paste0(mtt_results$concentration, " microM")
mtt_results$concentration <- factor(
  mtt_results$concentration,
  levels = c("1 microM", "5 microM", "10 microM")
)

mtt_results$pct_viability <- as.numeric(
  sub("%", "", mtt_results$pct_viability)
) / 100
```

```{r save the formatted results}
write_csv(
  mtt_results,
  here("MTT_Assay", "mtt_assay_results_formatted.csv")
)
```

# Visualize Biological Replicates
## All Replicates
```{r}
batch_df <- mtt_results |>
  mutate(group = paste0(cell_line, "_", timepoint)) |>
  relocate(group, .after = timepoint) |>
  group_by(group) |>
  arrange(cell_line, concentration) |> 
  ungroup()
```

```{r}
plot_list <- list()

for (i in unique(batch_df$group)) {
  # Extract group conditions
  line <- sapply(strsplit(i, "_"), `[`, 1)
  time <- sapply(strsplit(i, "_"), `[`, 2)
  
   # Filter the data for the current group
  group_df <- subset(batch_df, group == i)
  
  # Generate the plot
  p <- ggplot(group_df, aes(x = concentration, y = pct_viability, fill = date)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = paste0(line, " Treated with 13cRA for ", time),
    x = "Concentration of 13cRA",
    y = "Percent Viability",
    fill = "Experiment Date"
  ) +
  theme_minimal()
  
  # Add plot to plot list
  plot_list[[i]] <- p
}

pdf(
  here("MTT_Assay", "mtt_results_by_biological_replicates_all.pdf"),
  height = 12, width = 12
)

ggarrange(plotlist = plot_list, ncol = 3, nrow = 4)

dev.off()
```

# Remove the problematic plate
```{r}
mtt_results_filt <- mtt_results |>
  filter(date != "20241121")
```

# Reponse Across Cell Lines
```{r}
mean_results_filt <- mtt_results_filt |>
  group_by(cell_line, timepoint, concentration)|>
  mutate(group = paste0(cell_line, "_", timepoint, "_", concentration)) |>
  relocate(group, .after = timepoint) |>
  arrange(group) |>
  mutate(mean_pct_viability = mean(pct_viability)) |>
  ungroup()
```

```{r}
plot_list <- list()

for (i in levels(mean_results_filt$concentration)) {
  # Filter for how you want to split the data
  group_df <- subset(mean_results_filt, concentration = i)
  # Generate the plot
  p <- ggplot(
    mean_results_filt, 
    aes(x = factor(timepoint), y = mean_pct_viability, fill = cell_line)
  ) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(
      title = paste0("Cell Lines Treated with ", i, " 13cRA"),
      x = NULL,
      y = "Mean Percent Viability",
      fill = "Cell Line"
    ) +
  theme_minimal()
  
  # Add plot to plot list
  plot_list[[i]] <- p
}

pdf(
  here("MTT_Assay", "mtt_results_by_cell_line.pdf"),
  height = 3, width = 12
)

ggarrange(plotlist = plot_list, ncol = 3, nrow = 1)

dev.off()
```

```{r}
# Calculate mean and standard deviation
mean_results_filt <- mtt_results_filt |> 
  group_by(cell_line, timepoint, concentration) |> 
  mutate(
    group = paste0(cell_line, "_", timepoint, "_", concentration),
    mean_pct_viability = mean(pct_viability),
    sd_pct_viability = sd(pct_viability)  # Calculate standard deviation
  ) |> 
  relocate(group, .after = timepoint) |> 
  arrange(group) |> 
  ungroup()

# Plot list initialization
plot_list <- list()

# Order concentrations explicitly (1 μM, 5 μM, 10 μM)
ordered_concentrations <- c("1 microM", "5 microM", "10 microM")  # Replace with actual concentration levels if needed

for (i in ordered_concentrations) {
  # Filter for the current concentration
  group_df <- subset(mean_results_filt, concentration == i)
  
  # Generate the plot
  p <- ggplot(group_df, aes(x = factor(timepoint), y = mean_pct_viability, fill = cell_line)) +
    geom_bar(stat = "identity", position = "dodge", color = "#343434") + # Bar plot with black border
    geom_errorbar(
      aes(ymin = mean_pct_viability - sd_pct_viability, ymax = mean_pct_viability + sd_pct_viability),
      position = position_dodge(width = 0.9), width = 0.25, show.legend = FALSE
    ) + # Add error bars without legend
    geom_point(
      aes(y = pct_viability), 
      position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9), 
      size = 2, 
      color = "dimgray"
    ) + # Add points for individual replicates in dark gray
    labs(
      title = paste0("Cell Lines Treated with ", i, " 13cRA"),
      x = NULL,
      y = "Mean Percent Viability",
      fill = "Cell Line"
    ) +
    theme_minimal() +
    theme(legend.position = "right") # Adjust legend position
  
  # Add plot to plot list
  plot_list[[i]] <- p
}

# Save all plots to a PDF
pdf(here("MTT_Assay", "mtt_results_by_cell_line_error_bars.pdf"), height = 3, width = 12)
ggarrange(plotlist = plot_list, ncol = 3, nrow = 1)
dev.off()
```

```{r}
# Calculate mean and standard deviation
mean_results_filt <- mtt_results_filt |> 
  group_by(phenotype, timepoint, concentration) |> 
  mutate(
    group = paste0(phenotype, "_", timepoint, "_", concentration),
    mean_pct_viability = mean(pct_viability),
    sd_pct_viability = sd(pct_viability)  # Calculate standard deviation
  ) |> 
  relocate(group, .after = timepoint) |> 
  arrange(group) |> 
  ungroup()

# Plot list initialization
plot_list <- list()

# Order concentrations explicitly (1 μM, 5 μM, 10 μM)
ordered_concentrations <- c("1 microM", "5 microM", "10 microM")  # Replace with actual concentration levels if needed

for (i in ordered_concentrations) {
  # Filter for the current concentration
  group_df <- subset(mean_results_filt, concentration == i)
  
  # Generate the plot
  p <- ggplot(group_df, aes(x = factor(timepoint), y = mean_pct_viability, fill = phenotype)) +
    geom_bar(stat = "identity", position = "dodge", color = "#343434") + # Bar plot with black border
    geom_errorbar(
      aes(ymin = mean_pct_viability - sd_pct_viability, ymax = mean_pct_viability + sd_pct_viability),
      position = position_dodge(width = 0.9), width = 0.25, show.legend = FALSE
    ) + # Add error bars without legend
    geom_point(
      aes(y = pct_viability), 
      position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9), 
      size = 2, 
      color = "dimgray"
    ) + # Add points for individual replicates in dark gray
    labs(
      title = paste0("Phenotypes Treated with ", i, " 13cRA"),
      x = NULL,
      y = "Mean Percent Viability",
      fill = "Phenotype"
    ) +
    theme_minimal() +
    theme(legend.position = "right") # Adjust legend position
  
  # Add plot to plot list
  plot_list[[i]] <- p
}

# Save all plots to a PDF
pdf(here("MTT_Assay", "mtt_results_by_phenotype_error_bars.pdf"), height = 3, width = 12)
ggarrange(plotlist = plot_list, ncol = 3, nrow = 1)
dev.off()
```

# Response Across Phenotype
```{r}
mean_phenotype_filt <- mtt_results_filt |>
  group_by(phenotype, timepoint, concentration)|>
  mutate(group = paste0(phenotype, "_", timepoint, "_", concentration)) |>
  relocate(group, .after = timepoint) |>
  arrange(group) |>
  mutate(mean_pct_viability = mean(pct_viability)) |>
  ungroup()
```

```{r}
plot_list <- list()

for (i in levels(mean_phenotype_filt$concentration)) {
  # Filter for how you want to split the data
  group_df <- subset(mean_phenotype_filt, concentration = i)
  # Generate the plot
  p <- ggplot(
    mean_phenotype_filt, 
    aes(x = factor(timepoint), y = mean_pct_viability, fill = phenotype)
  ) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(
      title = paste0("Neuroblastoma Phenotypes Treated with ", i, " 13cRA"),
      x = NULL,
      y = "Mean Percent Viability",
      fill = "Phenotype"
    ) +
  theme_minimal()
  
  # Add plot to plot list
  plot_list[[i]] <- p
}

pdf(
  here("MTT_Assay", "mtt_results_by_phenotype.pdf"),
  height = 9, width = 6
)

ggarrange(plotlist = plot_list, ncol = 1, nrow = 3)

dev.off()
```

## Average Results by Replicate
### With 20241121
```{r}
avg_batch_df <- batch_df |>
  group_by(cell_line, timepoint, concentration) |>
  mutate(group = paste0(cell_line, "_", timepoint, "_", concentration)) |>
  arrange(group, concentration) |>
  mutate(mean_pct_viability = mean(pct_viability)) |>
  ungroup()

```

```{r}
plot_list <- list()

for (i in unique(avg_batch_df$group)) {
   # Filter the data for the current group
  group_df <- subset(avg_batch_df, group == i)
  
  # Generate the plot
  p <- ggplot(group_df, aes(x = timepoint, y = avg_pct_viability, fill = concentration)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ cell_line) +
  labs(
    x = "Timepoint",
    y = "Percent Viability",
    fill = "Concentration"
  ) +
  theme_minimal()
  
  # Add plot to plot list
  plot_list[[i]] <- p
}

pdf(
  here("MTT_Assay", "mtt_results_by_biological_replicates_avg.pdf"),
  height = 12, width = 12
)

ggarrange(plotlist = plot_list, ncol = 1, nrow = 4)

dev.off()
```

# End of script
```{r style code}
script <- rstudioapi::getSourceEditorContext()$path
style_file(script)
lint(script)
```

```{r clean environment}
rm(list = ls())
gc()
```
