---
title: "Correlation by Region"
author: "Victoria Flanary"
date: "2025-01-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objective
Correlate ATAC-seq peaks by region (promoter vs distal peaks).
Run hierarchical clustering on peaks.

# Set-up
```{r load packages}
library(tidyverse)
library(matrixStats)
library(ComplexHeatmap)
library(circlize)
library(rstudioapi)
library(here)
library(styler)
library(lintr)
```

```{r load data}
distal_fpkm <- read_csv(
  here("Sen_13cRA_ATAC", "data", "counts", "distal_regions_fpkm_counts.csv")
)

promoter_fpkm <- read_csv(
  here("Sen_13cRA_ATAC", "data", "counts", "promoters_fpkm_counts.csv")
)
```

# Format the data
```{r rm fragment col and convert to matrix}
distal_fpkm <- distal_fpkm[, -1] |> data.matrix()
promoter_fpkm <- promoter_fpkm[, -1] |> data.matrix()
```

# Correlation matrices
```{r make a function to generate correlation matrices}
corr_matrix <- function(counts_matrix) {
  # filter by row means and variance
  vars <- rowVars(counts_matrix)
  means <- rowMeans(counts_matrix)
  filt <- subset(counts_matrix, vars > 0 & means >= 1)

  # calculate correlation matrix
  cor(filt)
}
```

```{r get correlation matrices}
distal_corr <- corr_matrix(distal_fpkm)
promoter_corr <- corr_matrix(promoter_fpkm)
```

```{r visualize correlation matrices}
colors <- colorRamp2(c(-1, 0, 1), c("royalblue3", "white", "tomato2"))

pdf(
  here("Sen_13cRA_ATAC", "analysis", "correlate_peaks_by_region",
       "heatmaps_of_correlation_matrices.pdf"),
  height = 6, width = 6
)

Heatmap(promoter_corr,
  column_title = "Pearson Correlation of Promoter Chromatin Accessibility",
  name = "correlation", col = colors,
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2"
)

Heatmap(distal_corr,
  column_title = "Pearson Correlation of Distal and
        Intergenic Region Chromatin Accessibility",
  name = "correlation", col = colors,
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2"
)

dev.off()
```

# Hierarchical Clustering
```{r calculate euclidean distance}
euclidean <- dist(distal_corr, method = "euclidean")
```

```{r plot clusters}
pdf(
  here("Sen_13cRA_ATAC", "analysis", "correlate_peaks_by_region",
       "hierarchical_clustering_distal_peaks.pdf"),
  height = 6, width = 6
)

hc <- hclust(euclidean, method = "ward.D2")
plot(hc, main = "Hierarchichal Clustering of Distal Peaks")

dev.off()
```

```{r calculate euclidean distance}
euclidean <- dist(promoter_corr, method = "euclidean")
```

```{r plot clusters}
pdf(
  here("Sen_13cRA_ATAC", "analysis", "correlate_peaks_by_region",
       "hierarchical_clustering_promoter_peaks.pdf"),
  height = 6, width = 8
)

hc <- hclust(euclidean, method = "ward.D2")
plot(hc, main = "Hierarchichal Clustering of Promoter Peaks")

dev.off()
```

Both analyses show that promoter peaks cluster by technical artefacts, while
distal peaks cluster by known neuroblastoma phenotype. Proceed with only the
distal peaks for downstream analyses.

# End of script
```{r style code}
script <- rstudioapi::getSourceEditorContext()$path
style_file(script)
lint(script)
```

```{r clean environment}
rm(list = ls())
gc()
```