---
title: "Differential Peak Analysis"
author: "Victoria Flanary"
date: "2025-01-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objective
Run differential peak analysis for 13cRA-treated neuroblastoma lines.

# Set-up
```{r load packages}
library(DESeq2)
library(tidyverse)
library(matrixStats)
library(stats)
library(rstudioapi)
library(here)
library(styler)
library(lintr)
```

```{r load data}
counts <- read_csv(
  here("Sen_13cRA_ATAC", "data", "counts", "distal_regions_raw_counts.csv")
) |> as.data.frame()

coldata <- read.table(
  here("Sen_13cRA_ATAC", "data", "coldata.txt")
)
```

# Format the Data
```{r move fragment column to rownames}
rownames(counts) <- counts$fragments
counts <- counts[, colnames(counts) != "fragments"]
```

```{r ensure rownames in coldata correspond to counts cols}
counts <- counts[, match(rownames(coldata), colnames(counts))]
identical(rownames(coldata), colnames(counts)) # TRUE
```

# Filter Counts by Row Mean and Variance
```{r}
cts <- counts |>
  mutate(
    row_var = rowVars(data.matrix(counts)),
    row_mean = rowMeans(data.matrix(counts))
  ) |>
  filter(
    row_var > 0 & row_mean >= 10
  )

cts <- cts[, colnames(cts) %in% rownames(coldata)]
```

# Format colData
```{r convert all columns to factors}
str(coldata)
coldata <- as.data.frame(lapply(coldata, as.factor))
str(coldata)
```

# Case = MES
```{r set case and control}
coldata$group <- "control"
coldata$group[coldata$phenotype == "MES"] <- "case"
```

```{r generate the deseq object}
dds <- DESeqDataSetFromMatrix(
  countData = as.matrix(cts),
  colData = coldata,
  design = ~ mycn_status + group
)
```

Cannot include both age_months and sex/mycn_status (latter form the same groups)
in the design formula. Get the following error:
Error in checkFullRank(modelMatrix) : 
  the model matrix is not full rank, so the model cannot be fit as specified.
  One or more variables or interaction terms in the design formula are linear
  combinations of the others and must be removed.

  Please read the vignette section 'Model matrix not full rank':

  vignette('DESeq2')
In addition: Warning message:
In DESeqDataSet(se, design = design, ignoreRank) :
  some variables in design formula are characters, converting to factors
  
*Only include mycn_status as a covariate, since it's the most biologically relevant*

```{r relevel}
dds$group <- relevel(dds$group, "control")
```

```{r run the dea}
dds <- DESeq(dds, test = "Wald")
```

```{r annotate results in a df}
res <- as.data.frame(results(dds, contrast = c("group", "case", "control")))
peak_id <- rownames(res)
res <- cbind(peak_id, res)
rownames(res) <- NULL
```

```{r split the results by significance and direction}
sig <- res[which(res$padj < 0.05 & abs(res$log2FoldChange) >= 1.5), ]
nonsig <- res[which(res$padj >= 0.05 & abs(res$log2FoldChange) < 1.5), ]

# annotate significant differential peaks by up and down
sig$reg <- "down"
sig$reg[sig$log2FoldChange >= 1.5] <- "up"
```

```{r count how many differential peaks are up vs down}
tally <- sig |>
  group_by(reg) |>
  tally() |>
  data.frame()
```

```{r save differential results}
write_csv(
  res,
  here(
    "Sen_13cRA_ATAC", "analysis", "diff_peak_analysis",
    "mes_deseq2_all.csv"
  )
)

write_csv(
  sig,
  here(
    "Sen_13cRA_ATAC", "analysis", "diff_peak_analysis",
    "mes_deseq2_sig.csv"
  )
)

write_csv(
  tally,
  here(
    "Sen_13cRA_ATAC", "analysis", "diff_peak_analysis",
    "mes_deseq2_tally.csv"
  )
)
```

```{r identify top peaks}
head_genes <- sig |>
  arrange(desc(log2FoldChange)) |>
  slice_head(n = 10)

tail_genes <- sig |>
  arrange(log2FoldChange) |>
  slice_head(n = 10)

top_genes <- bind_rows(head_genes, tail_genes)
```

```{r merge results back together}
sig$significance <- "Significant"
nonsig$significance <- "Not Significant"

mes_results <- bind_rows(sig, nonsig)
```

```{r create the volcano plot}
library(ggrepel)

pdf(
  here("Sen_13cRA_ATAC", "analysis", "diff_peak_analysis", "mes_volcano_plot.pdf"),
  height = 7, width = 9
)

volcano_plot <- ggplot(mes_results, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = significance), alpha = 0.5, size = 2) +
  scale_color_manual(values = c("Significant" = "royalblue3", "Not Significant" = "gray")) +
  labs(
    title = "Peaks Enriched in MES Cells after 13cRA",
    x = "Log2 Fold Change",
    y = "-Log10 Adjusted P-value",
    color = "Significance"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",            # Keep the legend on the right
    plot.title = element_text(hjust = 0.5)  # Center the title
  )

print(volcano_plot)

dev.off()
```

# Case = ADR
```{r set case and control}
coldata$group <- "control"
coldata$group[coldata$phenotype == "ADR"] <- "case"
```

```{r generate the deseq object}
dds <- DESeqDataSetFromMatrix(
  countData = as.matrix(cts),
  colData = coldata,
  design = ~ mycn_status + group
)
```

```{r relevel}
dds$group <- relevel(dds$group, "control")
```

```{r run the dea}
dds <- DESeq(dds, test = "Wald")
```

```{r annotate results in a df}
res <- as.data.frame(results(dds, contrast = c("group", "case", "control")))
peak_id <- rownames(res)
res <- cbind(peak_id, res)
rownames(res) <- NULL
```

```{r split the results by significance and direction}
sig <- res[which(res$padj <= 0.05 & abs(res$log2FoldChange) >= 1.5), ]
nonsig <- res[which(res$padj >= 0.05 & abs(res$log2FoldChange) < 1.5), ] 

# annotate significant differential peaks by up and down
sig$reg <- "down"
sig$reg[sig$log2FoldChange >= 1.5] <- "up"
```

```{r count how many differential peaks are up vs down}
tally <- sig |>
  group_by(reg) |>
  tally() |>
  data.frame()
```

```{r save differential results}
write_csv(
  res,
  here(
    "Sen_13cRA_ATAC", "analysis", "diff_peak_analysis",
    "adr_deseq2_all.csv"
  )
)

write_csv(
  sig,
  here(
    "Sen_13cRA_ATAC", "analysis", "diff_peak_analysis",
    "adr_deseq2_sig.csv"
  )
)

write_csv(
  tally,
  here(
    "Sen_13cRA_ATAC", "analysis", "diff_peak_analysis",
    "adr_deseq2_tally.csv"
  )
)
```

```{r identify top peaks}
head_genes <- sig |>
  arrange(desc(log2FoldChange)) |>
  slice_head(n = 10)

tail_genes <- sig |>
  arrange(log2FoldChange) |>
  slice_head(n = 10)

top_genes <- bind_rows(head_genes, tail_genes)
```

```{r merge results back together}
sig$significance <- "Significant"
nonsig$significance <- "Not Significant"

adr_results <- bind_rows(sig, nonsig)
```

```{r create the volcano plot}
library(ggrepel)

pdf(
  here("Sen_13cRA_ATAC", "analysis", "diff_peak_analysis", "adr_volcano_plot.pdf"),
  height = 7, width = 9
)

volcano_plot <- ggplot(adr_results, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = significance), alpha = 0.5, size = 2) +
  scale_color_manual(values = c("Significant" = "chartreuse4", "Not Significant" = "gray")) +
  labs(
    title = "Peaks Enriched in ADR Cells after 13cRA",
    x = "Log2 Fold Change",
    y = "-Log10 Adjusted P-value",
    color = "Significance"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",            # Keep the legend on the right
    plot.title = element_text(hjust = 0.5)  # Center the title
  )

print(volcano_plot)

dev.off()
```
Should be exactly the same results, just in opposite directions.

# End of script
```{r style code}
script <- rstudioapi::getSourceEditorContext()$path
style_file(script)
lint(script)
```

```{r clean environment}
rm(list = ls())
gc()
```
